% ============================================================================
% Epidemiological Methodology --- Community Health Indicators (CHI) Reports
% ============================================================================
% Compile:  pdflatex epidemiological_methodology.tex  (×2 for TOC)
%           Or upload to https://overleaf.com (free, instant compile)
% ============================================================================

\documentclass[11pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{tocloft}
\usepackage{caption}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!60!black,
    urlcolor=blue!60!black,
    citecolor=blue!60!black
}

% ── Styling ────────────────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small Community Health Indicators --- Epidemiological Methodology}
\fancyhead[R]{\small CHI Report Views}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\title{\textbf{Epidemiological Methodology}\\[4pt]
       \large Community Health Indicators (CHI) Report Views}
\author{%
    \small CHI Reporting Project\\
    \small Data Source: NMR.LEANHIS EMR Schema (Snowflake)
}
\date{\small \today}

\begin{document}
\maketitle
\thispagestyle{empty}

\begin{abstract}
\noindent
This document defines the epidemiological methods underpinning the Community Health Indicators (CHI) reporting system. The system is organized into two methodologically distinct modules built on a common patient-month analytical grain across five chronic conditions: Diabetes Mellitus (DM), Hypertension (HTN), Dyslipidemia (DLP), Obesity (OB), and Prediabetes (PREDIAB).

\begin{itemize}[leftmargin=*, itemsep=2pt]
    \item \textbf{Module 1 --- Epidemiological Surveillance} answers \textit{how much disease exists and how fast is it appearing?} using monthly screening coverage (DM/HTN/DLP/OB only), annual point prevalence (all 5 conditions including Prediabetes), and monthly incidence reports (all 5 conditions including Prediabetes). Base population: \textbf{at-risk} cohort (undiagnosed) for Incidence; \textbf{total eligible} cohort for Prevalence. Thresholds: \textit{screening} cut-points.
    \item \textbf{Module 2 --- Compliance \& Care Gap} answers \textit{of those with disease, how well are we managing them?} using four report families: annual disease control classification (per condition), quarterly follow-up completion (per condition), annual care-gap distribution (per condition), and the \emph{generic} High-Risk Patients report (parameterized across conditions via the \texttt{chi\_high\_risk\_factors} config table; for v1 only Prediabetes is populated). Base population: \textbf{prevalent} cohort (diagnosed). Thresholds: \textit{control} targets plus condition-specific risk-factor definitions.
\end{itemize}

We describe the study design, cohort definitions, at-risk population dynamics, screening and control thresholds, rate calculation formulas, health cluster stratification, and the assumptions and limitations of each measure. The document serves as the methodological reference for analysts, epidemiologists, and public health practitioners interpreting CHI report outputs.
\end{abstract}

\vspace{12pt}
\tableofcontents
\newpage

% ============================================================================
\section{Study Design}
% ============================================================================

\subsection{Design Type}
Retrospective cohort study using electronic medical record (EMR) data from the \texttt{NMR.LEANHIS} Snowflake schema.

\subsection{Observation Period}
One calendar year (January 1 -- December 31), configurable via the \texttt{CHI\_REPORTING.chi\_config} parameter table. Changing the report year requires updating a single row; all downstream views reference this configuration.

\subsection{Unit of Analysis}
The \textbf{patient-month}. Each eligible patient contributes up to 12 monthly observations. The patient-month is the analytical grain from which all 10 epidemiological measures --- screening, prevalence, incidence, prediabetes prevalence, prediabetes incidence (Module~1), and the generic high-risk patients report (any condition), control, quarterly care gap, annual care gap (Module~2) --- are derived.

\subsection{Data Sources}

\begin{table}[ht]
\centering\small
\caption{Source EMR tables and their role in the CHI pipeline}
\begin{tabular}{@{}p{0.28\textwidth} p{0.32\textwidth} p{0.32\textwidth}@{}}
\toprule
\textbf{Table} & \textbf{Content} & \textbf{Used By} \\
\midrule
\texttt{PATIENTS} & Demographics, vital status & Cohort eligibility \\
\texttt{PATIENTVISITS} & Encounter dates & Monthly visit detection \\
\texttt{LABRESULTS} / \texttt{LABRESULTS\_RESULTVALUES} & Laboratory test results & DM, DLP screening \& control \\
\texttt{OBSERVATIONS} / \texttt{OBSERVATIONS\_OBSERVATIONVALUES} & Clinical observations (vitals) & All four conditions \\
\texttt{DIAGNOSIS\_CODES} & ICD-10 diagnosis records & Cohort classification, incidence detection \\
\texttt{PHC\_ASSIGNMENT} & Health cluster assignment & Stratification dimension (both modules) \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================================
\section{Population Eligibility}
% ============================================================================

A patient is included in the \textbf{total population} for report year $Y$ if they meet \textbf{all} of the following criteria at January 1 of year $Y$:

\begin{enumerate}[label=\textbf{C\arabic*}, leftmargin=*]
    \item \textbf{Age}: $> 18$ years at January 1. Pediatric thresholds differ for most chronic conditions and are outside the scope of this reporting system.
    \item \textbf{National ID}: Present and non-empty in the EMR. Ensures patient identity is verified and excludes transient or tourist encounters.
    \item \textbf{Vital status}: Alive at January 1 (no death record with a date before the report start). Deceased patients cannot contribute person-time during the report year.
\end{enumerate}

Patients meeting all three criteria are flagged $\text{is\_in\_total\_population} = \text{TRUE}$. This flag is evaluated once per report year and does \textit{not} change month-to-month. A patient who dies during the year remains in the denominator for that year's prevalence calculation (consistent with standard point prevalence methodology).

% ============================================================================
\section{Cohort Definitions}
% ============================================================================

From the total population, two mutually exclusive cohorts are derived based on diagnosis history. These cohorts underpin both modules of the CHI reporting system:

\begin{center}
\begin{tabular}{c}
\textbf{TOTAL POPULATION} (eligible at Jan 1) \\
$\downarrow$ \\
\begin{tabular}{p{0.42\textwidth} p{0.42\textwidth}}
\quad \textbf{PREVALENT COHORT} & \quad \textbf{AT-RISK COHORT} \\
\quad $\geq 1$ diagnosis in the condition's & \quad No diagnosis in the condition's \\
\quad ICD-10 code set (any date) & \quad ICD-10 code set \\
\quad $\rightarrow$ Module 2 base population & \quad $\rightarrow$ Module 1 screening \\
\quad \quad (Reports 4--6) & \quad \quad and incidence denominator \\
\quad $\rightarrow$ Module 1 prevalence numerator & \\
\end{tabular}
\end{tabular}
\end{center}

\subsection{Prevalent Cohort}

A patient is \textbf{prevalent} if they have \textit{any} ICD-10 code in the condition's code set, regardless of whether it is the primary target code. This cohort is the base population for:

\begin{itemize}[leftmargin=*]
    \item \textbf{Module 1} --- Report 2 (Prevalence numerator)
    \item \textbf{Module 2} --- Reports 4, 5, 6 (denominator of all care-gap and control metrics)
\end{itemize}

\begin{table}[ht]
\centering
\caption{ICD-10 code sets and target codes by condition}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Condition} & \textbf{ICD-10 Code Set (Prevalent)} & \textbf{Target Code (Incidence)} \\
\midrule
Diabetes Mellitus (DM) & E10, E11, E13, E14, O24 & E11 \\
Hypertension (HTN) & I10, I11, I12, I13, I15 & I10 \\
Dyslipidemia (DLP) & E78 & E78 \\
Obesity (OB) & E66 & E66 \\
Prediabetes (PREDIAB) & R73.03 & R73.03 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Rationale for broader code set in DM.}
A patient with Type 1 diabetes (E10) or gestational diabetes (O24) is already a known diabetic---screening them for Type 2 DM makes no clinical sense. The at-risk pool should exclude all known diabetics. However, incidence tracking focuses specifically on Type 2 (E11) because it is the preventable, lifestyle-related form that public health programs target.

\subsection{At-Risk Cohort}

\begin{equation}
\text{AT-RISK} = \text{TOTAL POPULATION} - \text{PREVALENT COHORT}
\end{equation}

The at-risk cohort is used \textbf{only by Module 1} (Reports 1 and 3) and includes:
\begin{itemize}[leftmargin=*]
    \item Patients with \textbf{no diagnosis} in the condition's code set before the report year
    \item Patients with \textbf{prediabetes} or abnormal screening results who lack a formal ICD-10 diagnosis
    \item Patients who have \textbf{never been screened}
\end{itemize}

\paragraph{Prediabetes patients and the DM at-risk pool.}
A patient with a prediabetes ICD-10 code (R73.03) is, by definition, a known diabetic-risk patient with no formal DM diagnosis. They are therefore included in the DM at-risk pool (they can be screened and can become an incident E11 case). However, for the \textbf{Prediabetes Incidence report} (§6.3), the at-risk pool is \textbf{inverted}: prediabetes patients (R73.03 prevalent) are \emph{excluded} from the prediabetes at-risk pool, because they are the \emph{prevalent} cohort for the report. The two cohorts (\texttt{is\_in\_at\_risk} for DM reports vs.\ \texttt{is\_in\_at\_risk\_prediab} for the Prediabetes Incidence report) are deliberately inverse.

\subsection{Dynamic Cohort Membership}

Cohort assignment is \textbf{not static} across the year. The at-risk pool shrinks month-by-month as patients receive their first diagnosis and transition to the prevalent cohort. The staging system evaluates $\text{is\_at\_risk\_start}$ at each month boundary, not once at the start of the year. This dynamic evaluation is relevant only to Module 1; Module 2's base population is fixed at a single reference date (see §7.0).

\begin{figure}[ht]
\centering
\begin{tabular}{ccccccccccccccc}
 & & \multicolumn{2}{c}{$\xleftarrow{\text{Incident in Mar}}$} \\
Jan & -- & Feb & -- & Mar & -- & Apr & -- & May & -- & Jun & -- & $\cdots$ & -- & Dec \\[6pt]
\multicolumn{15}{c}{At-risk at Jan 1: 13 patients \hspace{2cm} At-risk at Dec 1: 10 patients}
\end{tabular}
\caption{Dynamic at-risk population across the report year. A patient diagnosed in March leaves the at-risk pool for April--December.}
\end{figure}

% ============================================================================
\section{The At-Risk Population}
% ============================================================================

\subsection{Definition}

For month $M$:

\begin{equation}
\text{is\_at\_risk\_start}(M) =
\begin{cases}
\text{TRUE} & \text{if patient is in TOTAL POPULATION} \\
            & \text{AND has no diagnosis in the condition's ICD-10 set} \\
            & \text{with diagnosis date} < \text{first day of month } M \\[4pt]
\text{FALSE} & \text{otherwise}
\end{cases}
\end{equation}

\subsection{Why Monthly Evaluation?}

Three epidemiological reasons:

\begin{enumerate}[leftmargin=*]
    \item \textbf{Accurate denominators}. The at-risk population is the denominator for both screening and incidence rates. Using a fixed year-start denominator would overstate the population at risk in later months.
    \item \textbf{Censoring at diagnosis}. A patient can only be an incident case once. After their first diagnosis, they leave the at-risk pool. The monthly evaluation correctly censors them from subsequent months' denominators.
    \item \textbf{Time-varying exposure}. Screening behavior changes when patients know they have a condition. A patient diagnosed in March should not be counted as ``at-risk but unscreened'' in November---they are already under management.
\end{enumerate}

% ============================================================================
\section{Screening Thresholds}
% ============================================================================

Screening thresholds classify \textbf{at-risk} patients (Module 1 base population) into Normal / Elevated / Abnormal categories based on the most recent screening test result. These cut-points are used by Module 1's Screening report (Report 1) only. Module 2 uses a separate set of control thresholds (§7.5) that classify \textit{prevalent} patients into control tiers.

\subsection{Diabetes Mellitus}

Screening category = $\max(\text{FBS\_category}, \text{A1C\_category})$. FBS auto-detects units by value range ($< 30$ = mmol/L, $\geq 30$ = mg/dL).

\begin{table}[ht]
\centering
\caption{DM screening thresholds}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Category} & \textbf{FBS (mmol/L)} & \textbf{FBS (mg/dL)} & \textbf{HbA1c (\%)} \\
\midrule
Normal   & $\leq 5.5$    & $\leq 99$    & $< 5.7$ \\
Elevated & $5.6$--$6.9$  & $100$--$125$ & $5.7$--$6.4$ \\
Abnormal & $> 6.9$       & $> 125$      & $> 6.4$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Hypertension (ACC/AHA 2017)}

SYS and DIA must be \textbf{paired from the same visit}. Classification = worst of SYS or DIA.

\begin{table}[ht]
\centering
\caption{HTN screening thresholds}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Category} & \textbf{Systolic (mmHg)} & \textbf{Diastolic (mmHg)} \\
\midrule
Normal   & $< 120$ \textbf{and} $< 80$ \\
Elevated & $120$--$129$ \textbf{or} $80$--$89$ \\
Abnormal & $\geq 130$ \textbf{or} $\geq 90$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Dyslipidemia}

Screening category = $\max(\text{HDL}, \text{Triglyceride}, \text{Cholesterol}, \text{LDL})$.

\textbf{HDL} (gender-specific; no ``elevated'' tier):

\begin{table}[ht]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Gender} & \textbf{Normal} & \textbf{Abnormal} \\
\midrule
Male   & $\geq 40$ & $< 40$ \\
Female & $\geq 50$ & $< 50$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Triglyceride, Cholesterol, LDL}:

\begin{table}[ht]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Marker} & \textbf{Normal} & \textbf{Elevated} & \textbf{Abnormal} \\
\midrule
Triglyceride      & $< 150$ & $150$--$199$ & $\geq 200$ \\
Total Cholesterol & $< 200$ & $200$--$239$ & $\geq 240$ \\
LDL               & $< 130$ & $130$--$159$ & $\geq 160$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Obesity (WHO BMI)}

\begin{table}[ht]
\centering
\caption{Obesity screening thresholds}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Category} & \textbf{BMI (kg/m}$^2$\textbf{)} \\
\midrule
Underweight           & $< 18.5$ \\
Normal                & $18.5$--$24.9$ \\
Elevated (Overweight) & $25.0$--$29.9$ \\
Abnormal (Obese)      & $\geq 30.0$ \\
\bottomrule
\end{tabular}
\end{table}

BMI values $< 10$ or $> 80$ are excluded as clinically implausible (likely data entry errors).

% ============================================================================
\section{Module 1: Epidemiological Surveillance}
% ============================================================================

Module 1 answers: \textit{How much disease exists in the population, how fast is it appearing, and are we screening effectively?} Three reports derive from the \textbf{at-risk} cohort (§3.2), using \textbf{screening thresholds} (§5).

% --------------------------------------------------------------------------
\subsection{Report 1 --- Screening (Monthly Coverage)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
\textit{What proportion of the at-risk population received screening for the condition each month?}

\subsubsection{Measure Type}
\textbf{Monthly screening coverage rate}---a process measure reflecting healthcare system reach.

\subsubsection{Formula}

\begin{equation}
\text{Screening Rate}_{M} =
\frac{\text{Patients screened during month } M}
     {\text{Patients at-risk at start of month } M} \times 100
\end{equation}

\subsubsection{Numerator}

A patient is counted as \textbf{screened} in month $M$ if:
\begin{enumerate}[leftmargin=*]
    \item They are at-risk at the start of month $M$ ($\text{is\_at\_risk\_start} = \text{TRUE}$), \textbf{and}
    \item They had the required screening test(s) recorded during month $M$.
\end{enumerate}

\begin{table}[ht]
\centering
\caption{Screening requirements by condition}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Condition} & \textbf{Required Test(s)} & \textbf{Logic} \\
\midrule
DM & Fasting Blood Sugar (FBS) \textbf{or} HbA1c & Either test counts \\
HTN & Systolic BP \textbf{and} Diastolic BP (same visit) & Both must be present \\
DLP & HDL \textbf{or} LDL & Either lipid component counts \\
OB & BMI & Single measurement \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Denominator}

The at-risk population at the start of month $M$.

\subsubsection{Stratification by Result Category}

Screened patients are classified by their \textbf{worst} result (the most clinically concerning marker determines the category) using the screening thresholds in §5:

\begin{table}[ht]
\centering
\caption{Screening result categories}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Category} & \textbf{Clinical Meaning} \\
\midrule
Normal & All markers within normal range \\
Elevated & $\geq 1$ marker in borderline range, none abnormal \\
Abnormal & $\geq 1$ marker in disease-indicating range \\
\bottomrule
\end{tabular}
\end{table}

For Obesity, a fourth category exists: \textbf{Underweight} (BMI $< 18.5$).

\subsubsection{Interpretation Guide}

\begin{itemize}[leftmargin=*]
    \item \textbf{High screening rate + low abnormal rate}: Effective prevention; population is healthy.
    \item \textbf{High screening rate + high abnormal rate}: Good detection; high undiagnosed burden exists.
    \item \textbf{Low screening rate}: Insufficient healthcare reach; disease burden is unknown.
    \item \textbf{Zero screening}: No data; the at-risk population's status is unknown.
\end{itemize}

\subsubsection{Annual Cumulative Rate (Subtotal Rows)}

\begin{equation}
\text{Cumulative Screening Rate}_{\text{annual}} =
\frac{\sum_{m=1}^{12} \text{Screened}_m}
     {\sum_{m=1}^{12} \text{At-Risk}_m} \times 100
\end{equation}

This is a \textbf{person-month rate}: a patient screened in 3 different months contributes 3 to the numerator and 3 person-months to the denominator. It measures screening \textit{encounters}, not unique patients screened per year.

% --------------------------------------------------------------------------
\subsection{Report 2 --- Prevalence (Annual Point Prevalence)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
\textit{What proportion of the total eligible population had the condition at the end of the report year?}

\subsubsection{Measure Type}
\textbf{Point prevalence}---a cross-sectional snapshot as of December 31.

\subsubsection{Formula}

\begin{equation}
\text{Prevalence}_{\text{annual}} =
\frac{\text{Patients with target ICD-10 code at Dec 31}}
     {\text{Total eligible population at Jan 1}} \times 100
\end{equation}

\subsubsection{Numerator}

A patient is counted as \textbf{prevalent} if they have the target ICD-10 code with a diagnosis date $\leq$ December 31 of the report year.

\subsubsection{Denominator}

The total eligible population at January 1. This denominator is \textbf{fixed} for the year.

\paragraph{Rationale for fixed denominator.}
Point prevalence is a cross-sectional measure. The January 1 population represents the community at the start of the observation period. Removing patients who die during the year would underestimate prevalence.

\subsubsection{Sub-Components}

\begin{equation}
\text{Prevalent} = \text{Pre-existing} + \text{Incident During Year}
\end{equation}

\begin{table}[ht]
\centering
\caption{Prevalence sub-components}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Component} & \textbf{Definition} & \textbf{Clinical Meaning} \\
\midrule
Pre-existing & First diagnosis \textbf{before} Jan 1 & Known cases; already under management \\
Incident during year & First diagnosis \textbf{during} report year & Newly detected cases \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Interpretation}

\begin{itemize}[leftmargin=*]
    \item Rising prevalence over years may indicate improved detection, increasing disease burden, or both.
    \item A high ratio of pre-existing to incident cases suggests a mature detection program.
    \item A high ratio of incident to pre-existing cases suggests either a recent screening push or genuinely increasing disease incidence.
\end{itemize}

% --------------------------------------------------------------------------
\subsection{Report 3 --- Incidence (Monthly Incidence Rate)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
\textit{At what rate are new cases of the condition developing in the at-risk population?}

\subsubsection{Measure Type}
\textbf{Monthly cumulative incidence proportion} and \textbf{annualized incidence rate} (using January baseline).

\subsubsection{Monthly Formula}

\begin{equation}
\text{Incidence Rate}_{M} =
\frac{\text{New cases diagnosed during month } M}
     {\text{Patients at-risk at start of month } M} \times 100{,}000
\end{equation}

\subsubsection{Incident Case Definition}

A patient is an \textbf{incident case} in month $M$ if \textbf{all} of the following are true:
\begin{enumerate}[leftmargin=*]
    \item They are at-risk at the start of month $M$ (no prior diagnosis of the target condition).
    \item They received the \textbf{first-ever} target ICD-10 code during month $M$.
    \item The diagnosis date falls within the calendar boundaries of month $M$.
\end{enumerate}

A patient can be incident \textbf{only once}. After their first diagnosis, they are removed from the at-risk pool for all subsequent months.

\subsubsection{Why Per 100,000?}

Incidence rates are expressed per 100,000 population to produce readable numbers. Monthly incidence of chronic diseases is typically low (a few cases per thousand per month), so per-100,000 scaling avoids small decimals.

\subsubsection{Annual Incidence Rate (Subtotal Rows)}

\begin{equation}
\text{Annual Incidence Rate} =
\frac{\sum_{m=1}^{12} \text{Incident Cases}_m}
     {\text{At-Risk Population at January}} \times 100{,}000
\end{equation}

\paragraph{Rationale for January denominator.}
Using a fixed baseline (the population at risk on January 1) provides a stable denominator for comparison across clusters and years. Summing monthly denominators would make later months' cases appear more concentrated as the at-risk pool shrinks. The January baseline answers: \textit{Of the population at risk on January 1, how many developed the condition during the year?}

\paragraph{Per-condition target codes.}
\begin{table}[ht]
\centering\small
\caption{Per-condition code sets and at-risk pool semantics for the Incidence report}
\begin{tabular}{@{}p{0.20\textwidth} p{0.20\textwidth} p{0.16\textwidth} p{0.34\textwidth}@{}}
\toprule
\textbf{Condition} & \textbf{Prevalent Code Set} & \textbf{Target Code} & \textbf{At-Risk Pool Semantics} \\
\midrule
Diabetes Mellitus (DM) & E10, E11, E13, E14, O24 & E11 & Excludes DM patients; \textbf{includes} R73.03 prediabetes patients \\
Hypertension (HTN) & I10, I11, I12, I13, I15 & I10 & Excludes HTN patients \\
Dyslipidemia (DLP) & E78 & E78 & Excludes DLP patients \\
Obesity (OB) & E66 & E66 & Excludes OB patients \\
\textbf{Prediabetes (PREDIAB)} & \textbf{R73.03} & \textbf{R73.03} & \textbf{Excludes R73.03 patients (inverse of DM)} \\
\bottomrule
\end{tabular}
\end{table}

\textit{Prediabetes incidence} is structurally identical to the DM/HTN/DLP/OB incidence reports — same formula, same at-risk pool mechanics. The single difference is the inverse at-risk semantics: R73.03 patients are \emph{prevalent} (not at-risk) for this report, so they drop out of the denominator. See the callout in §3.2 for the rationale. SQL view: \texttt{rpt\_prediab\_incidence\_monthly}.

% --------------------------------------------------------------------------
\subsection{Monthly vs.\ Annual Rate Calculations}
% --------------------------------------------------------------------------

\begin{table}[ht]
\centering
\caption{Comparison of monthly and annual rate calculation methods}
\begin{tabular}{@{}p{0.22\textwidth} p{0.35\textwidth} p{0.35\textwidth}@{}}
\toprule
\textbf{Aspect} & \textbf{Monthly Rate} & \textbf{Annual Rate (Subtotal/Grand Total)} \\
\midrule
Denominator & At-risk at month start (varies each month) & January at-risk (fixed baseline) \\
Numerator & Cases/events in that specific month & Sum of all monthly cases/events \\
Interpretation & ``This month, $X$ per 100,000 at-risk developed the condition'' & ``Of those at risk on Jan 1, $X$ per 100,000 developed the condition during the year'' \\
Use case & Monitoring month-to-month variation, seasonal patterns & Comparing across clusters, facilities, or years \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Worked Example}

Using the DM simulation data:

\begin{table}[ht]
\centering
\caption{Illustration of monthly vs.\ annual incidence rates}
\begin{tabular}{@{}cccc@{}}
\toprule
\textbf{Month} & \textbf{At-Risk Start} & \textbf{New Cases} & \textbf{Monthly Rate (per 100k)} \\
\midrule
Jan  & 13 & 0 & 0 \\
Feb  & 13 & 1 & 7,692 \\
Mar  & 12 & 1 & 8,333 \\
Jun  & 11 & 1 & 9,091 \\
Dec  & 10 & 1 & 10,000 \\
\midrule
\textbf{Annual} & \textbf{13 (Jan)} & \textbf{4 (total)} & \textbf{30,769} \\
\bottomrule
\end{tabular}
\end{table}

The monthly rate rises across the year (0 $\rightarrow$ 10,000) even with a constant 1 case/month because the denominator shrinks. The annual rate (30,769 per 100k) uses the fixed January denominator of 13 to provide a single, comparable number.

% --------------------------------------------------------------------------
\subsection{Health Cluster Stratification}
% --------------------------------------------------------------------------

\subsubsection{Rationale}
Health clusters represent organizational units (geographic regions, facility catchments, or care networks). Stratifying epidemiological measures by cluster answers:
\begin{itemize}[leftmargin=*]
    \item Are some clusters screening more effectively than others?
    \item Is disease burden concentrated in specific clusters?
    \item Are new cases appearing at different rates across clusters?
    \item Which clusters have patients with no PHC assignment?
\end{itemize}

Cluster stratification is applied identically across all Module 1 and Module 2 reports; see §7 for the Module-2-specific rate forms.

\subsubsection{Method}
Health cluster is a \textbf{patient-level attribute} from \texttt{NMR.LEANHIS.PHC\_ASSIGNMENT}. It is attached at the cohort level and flows through to all reports:

\begin{equation}
\text{Metric}_{\text{cluster}} =
\frac{\text{Numerator}_{\text{cluster}}}
     {\text{Denominator}_{\text{cluster}}} \times \text{Scaling Factor}
\end{equation}

Each cluster's rate is calculated independently using that cluster's own numerator and denominator. The grand total row sums across all clusters.

\subsubsection{Unassigned Patients}
Patients with no PHC record are labeled `Unassigned' and reported as a separate group. This is methodologically important: if unassigned patients have systematically different screening rates or disease burden, it may indicate a data quality issue or a genuinely underserved population.


% ============================================================================
\section{Module 2: Compliance \& Care Gap}
% ============================================================================

Module 2 answers: \textit{Of those with disease, how well are we managing them and closing follow-up gaps?} Module 2 reports derive from the \textbf{prevalent} cohort (§3.1). Four report families:
\begin{itemize}[leftmargin=*]
    \item \textbf{Report 4} --- Control Level (Annual) --- disease control classification using §7.5
    \item \textbf{Report 5} --- Care Gap (Quarterly) --- follow-up completion rates
    \item \textbf{Report 6} --- Care Gap (Annual) --- annual distribution of follow-up completeness
    \item \textbf{Report 7} --- High-Risk Patients (Annual) --- \emph{generic}, parameterized by condition via \texttt{chi\_high\_risk\_factors}; for v1 only Prediabetes is populated. Covers all 5 conditions.
\end{itemize}

\subsection{Module 2 Preamble}

\paragraph{Base population.}
The base population for all Module 2 reports is the \textbf{prevalent cohort at the reference date}, as defined in §3.1. Unlike Module 1's at-risk pool, no further exclusion is applied for incident cases within the report year --- any patient diagnosed at any time before the reference date is included.

\paragraph{Reference date.}
\begin{itemize}[leftmargin=*]
    \item For \textbf{Report 4} (Control Level), \textbf{Report 6} (Care Gap Annual), and \textbf{Report 7} (High-Risk Patients): the prevalent cohort is evaluated as of the end of the report year (December 31).
    \item For \textbf{Report 5} (Care Gap Quarterly): the prevalent cohort is evaluated at the start of each quarter, so quarters-completed counts include all patients diagnosed before that quarter began.
\end{itemize}

\paragraph{Configuration.}
All control thresholds, the quarterly target, and the high-risk-factor definitions are stored in dedicated config tables --- \texttt{CHI\_REPORTING.chi\_control\_thresholds}, \texttt{CHI\_REPORTING.chi\_care\_gap\_config}, and \texttt{CHI\_REPORTING.chi\_high\_risk\_factors} --- rather than hardcoded in views. Updating thresholds or risk factors does not require view re-creation; only the config tables are updated.

\paragraph{Condition-specific follow-up markers.}
Each condition defines a marker that indicates a meaningful follow-up visit:

\begin{table}[ht]
\centering
\caption{Follow-up "completed" criteria by condition}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Condition} & \textbf{Completed if patient had ...} \\
\midrule
DM & $\geq 1$ HbA1c measurement that quarter \\
HTN & $\geq 1$ visit with paired SYS \textbf{and} DIA that quarter \\
DLP & $\geq 1$ lipid panel component (HDL/LDL/CHOL/TRIG) that quarter \\
OB  & $\geq 1$ BMI measurement that quarter \\
\bottomrule
\end{tabular}
\end{table}

% --------------------------------------------------------------------------
\subsection{Report 7 --- High-Risk Patients (Annual)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
Of patients diagnosed with a given condition by year-end, what proportion carry $\geq 2$ risk factors that signal imminent progression or complication --- and therefore warrant elevated clinical attention?

\subsubsection{Measure Type}
\textbf{Clinical risk stratification} --- an annual prevalence-of-risk-factor-burden snapshot per condition. The denominator is the \textbf{prevalent cohort at year-end} (Module 2 convention, like Reports 4--6), and the numerator is the subset of prevalent patients whose risk-factor count exceeds the configurable threshold.

This is a \emph{generic} report: the underlying view (\texttt{rpt\_high\_risk\_patients\_annual}) runs across all five conditions in one pass, parameterized by the \texttt{chi\_high\_risk\_factors} config table. For v1 only Prediabetes is populated, but the report is designed to extend to any condition with one SQL-only change (INSERT more config rows).

\subsubsection{Definition: High-Risk}
A patient is \textbf{high-risk} for condition \textit{C} if, at year-end:
\begin{enumerate}[leftmargin=*]
    \item They carry the condition's target ICD-10 code (prevalent at year-end) \textbf{AND}
    \item They have \textbf{$\geq N$ weighted risk factors} (default $N = 2$) drawn from condition-specific rows in \texttt{chi\_high\_risk\_factors}.
\end{enumerate}

For v1 Prediabetes, the 6 risk factors are documented in the new §7.1.1 below.

\paragraph{§7.1.1: Prediabetes risk factors (v1).}
\begin{table}[ht]
\centering\small
\begin{tabular}{@{}p{0.32\textwidth} p{0.42\textwidth} p{0.18\textwidth}@{}}
\toprule
\textbf{Factor} & \textbf{Source} & \textbf{Detail} \\
\midrule
BMI $\geq$ 25 & \texttt{stg\_prediab\_cohort.has\_bmi\_ge\_25} & Latest 2025 BMI reading (10--80 range) \\
Hypertension dx & \texttt{stg\_prediab\_cohort.has\_htn\_dx} & Reuses \texttt{stg\_htn\_cohort.has\_any\_htn\_diagnosis} \\
Dyslipidemia dx & \texttt{stg\_prediab\_cohort.has\_dlp\_dx} & Reuses \texttt{stg\_dlp\_cohort.has\_any\_dlp\_diagnosis} \\
First-degree family hx of DM & \emph{(sentinel)} & \textbf{Hardcoded FALSE} --- placeholder until structured family-history source is wired in \\
Gestational DM history & \texttt{stg\_prediab\_cohort.has\_gdm\_history} & Reuses \texttt{stg\_dm\_cohort.has\_gdm} (O24) \\
PCOS / PMOS & \texttt{stg\_prediab\_cohort.has\_pcos} & ICD-10 E28.2 (polycystic ovary syndrome) \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Base Population (per condition)}
For each condition, the denominator is \textbf{all patients whose target ICD-10 code is present at year-end} (\texttt{first\_target\_date < report\_end}). For Prediabetes: \texttt{first\_r73\_date < report\_end}. For DM: \texttt{first\_e11\_date < report\_end}. Etc.

\subsubsection{Configuration: chi\_high\_risk\_factors}
The risk-factor definitions live in a config table (mirrors the \texttt{chi\_control\_thresholds} pattern). Schema:

\begin{table}[ht]
\centering\small
\begin{tabular}{@{}p{0.22\textwidth} p{0.10\textwidth} p{0.60\textwidth}@{}}
\toprule
\textbf{Column} & \textbf{Type} & \textbf{Purpose} \\
\midrule
\texttt{condition} & VARCHAR & \texttt{'dm'} / \texttt{'htn'} / \texttt{'dlp'} / \texttt{'ob'} / \texttt{'prediab'} \\
\texttt{factor\_code} & VARCHAR & Short identifier (e.g.\ \texttt{'bmi\_ge\_25'}) \\
\texttt{factor\_label} & VARCHAR & Display name for the report \\
\texttt{source\_view} & VARCHAR & View/table to LEFT JOIN (e.g.\ \texttt{'CHI\_REPORTING.stg\_htn\_cohort'}) \\
\texttt{source\_column} & VARCHAR & Boolean column on that view; use \texttt{'always\_false'} sentinel for unimplemented factors \\
\texttt{value\_min} & DECIMAL(10,2) & Numeric threshold (BMI $\geq$ 25); NULL for flag-based factors \\
\texttt{weight} & INTEGER & Contribution to \texttt{risk\_factor\_count} (default 1) \\
\texttt{requires\_value} & BOOLEAN & TRUE = numeric threshold; FALSE = boolean flag \\
\texttt{level\_order} & INTEGER & Display order in the report \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Extending to a new condition} (e.g.\ adding High-Risk logic to DM):
\begin{enumerate}[leftmargin=*]
    \item INSERT new rows into \texttt{chi\_high\_risk\_factors} with \texttt{condition = 'dm'}
    \item Add a \texttt{CASE} branch in \texttt{stg\_high\_risk\_patient.factor\_evaluations} for the new \texttt{source\_view}/\texttt{source\_column} pair
    \item No view re-creation required --- the report auto-detects new conditions at query time
\end{enumerate}

For v1 only Prediabetes is wired through \texttt{stg\_high\_risk\_patient}. Other conditions return zero rows because their source columns are not yet referenced in the view's CASE chain.

\subsubsection{Generic Formula}
\begin{equation}
\text{High-Risk Pct}_{C, \text{cluster}} = \frac{\text{Prevalent patients of condition } C \text{ in cluster with risk\_factor\_count} \geq N}{\text{Prevalent patients of condition } C \text{ in cluster}} \times 100
\end{equation}

\begin{itemize}[leftmargin=*]
    \item \textbf{Numerator}: condition-\textit{C} prevalent patients where \texttt{is\_high\_risk = TRUE} (i.e.\ their \texttt{risk\_factor\_count} $\geq N$).
    \item \textbf{Denominator}: condition-\textit{C} prevalent patients at year-end.
\end{itemize}

\subsubsection{Report Structure}
Two-layer rows per condition: detail (\texttt{sort\_order=0}, per cluster) + grand total (\texttt{sort\_order=2}, \texttt{── ALL CLUSTERS ──}). For conditions with no factors defined, the section is omitted entirely (no zero-filled rows; the report only appears for populated conditions).

Output columns: \texttt{condition}, \texttt{health\_cluster}, \texttt{total\_prevalent}, \texttt{high\_risk\_count}, \texttt{high\_risk\_pct}, \texttt{sort\_key}, \texttt{sort\_order}.

\subsubsection{Interpretation Guide}
\begin{itemize}[leftmargin=*]
    \item \textbf{High HR \% in a cluster} (within a condition): that cluster's prevalent pool is dominated by the highest-concern cases. Clinical outreach and tighter follow-up should be prioritized there.
    \item \textbf{Low \texttt{total\_prevalent} but high HR \%}: under-diagnosis of mild cases in the cluster --- cases are surfacing only when they already carry multiple risk factors.
    \item \textbf{Large \texttt{total\_prevalent} but low HR \%}: detection is catching cases earlier; the prevalent pool is mostly low-risk.
    \item \textbf{Empty section for a condition}: no risk factors have been defined in the config. The report is silent rather than producing zero-filled rows. See §8.5 for caveats.
\end{itemize}

\subsubsection{Relationship to Reports 4--6}
Like Reports 4--6, Report 7 sits in Module 2 (it operates on the prevalent cohort). It differs in scope: Reports 4--6 are \emph{condition-specific} views repeated per condition, whereas Report 7 is a \emph{single generic view} parameterized across all five conditions. Conceptually it answers a different question (``who among our diagnosed patients carries the most clinical risk?'') from Reports 4--6 (``how well is disease being managed?''), but the denominator convention is identical --- and the config-driven join makes it the natural place for future cross-condition risk-profile extensions.

% --------------------------------------------------------------------------
\subsection{Report 4 --- Control Level (Annual)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
\textit{Of patients diagnosed with the condition, what proportion have their disease under control based on the most recent monitoring marker?}

\subsubsection{Measure Type}
\textbf{Annual disease control classification} --- an outcome proxy reflecting adequacy of disease management.

\subsubsection{Base Population}
\textbf{Prevalent cohort as of December 31} (§3.1). Each prevalent patient contributes one row.

\subsubsection{Formula --- Patient-Level Classification}

For each prevalent patient, the most recent monitoring marker value(s) within the report year are extracted and classified against the control thresholds (§7.5):

\begin{equation}
\text{level\_order}_{\text{patient}} = \max_{m \in \text{markers}} \left( \text{level\_order}( \text{value}_{m}, \text{thresholds}_{m} ) \right)
\end{equation}

\begin{itemize}[leftmargin=*]
    \item DM: single marker (A1C) --- $\text{level\_order}_{\text{patient}} = \text{level\_order}(\text{A1C})$
    \item HTN: two markers (SYS, DIA) from the most recent paired visit --- $\max(\text{SYS\_level}, \text{DIA\_level})$
    \item DLP: four markers (HDL, LDL, CHol, Trig) --- $\max(\text{HDL\_level}, \text{LDL\_level}, \text{CHol\_level}, \text{Trig\_level})$
    \item OB: single marker (BMI) --- $\text{level\_order}_{\text{patient}} = \text{level\_order}(\text{BMI})$
\end{itemize}

$\text{level\_order} = 0$ is best (controlled); higher values indicate worsening control. The control \textit{label} (the human-readable tier name) is looked up from the threshold row that matched the patient's value.

\subsubsection{Report-Level Formula}

\begin{equation}
\text{Pct}_{\text{level}, \text{cluster}} =
\frac{\text{Prevalent patients in control level (cluster)}}
     {\text{Prevalent patients (cluster)}} \times 100
\end{equation}

The aggregated report groups prevalent patients by \texttt{health\_cluster} $\times$ \texttt{control\_level}, with cluster subtotals and a grand total row.

\subsubsection{Stratification by Marker Set}

\begin{table}[ht]
\centering
\caption{Markers used for control classification by condition}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Condition} & \textbf{Markers} & \textbf{Aggregation} \\
\midrule
DM  & A1C only & Single-marker classification \\
HTN & SYS + DIA (paired) & $\max$ of two marker levels \\
DLP & HDL, LDL, CHol, Trig & $\max$ of four marker levels \\
OB  & BMI only & Single-marker classification \\
\bottomrule
\end{tabular}
\end{table}

HDL has gender-specific control thresholds (Male: $\geq 40$ controlled; Female: $\geq 50$ controlled) and is inverted --- lower HDL = worse control.

\subsubsection{Not Monitored}

Prevalent patients with no measurement of the relevant marker during the report year are classified as \textbf{Not Monitored}. This category is reported explicitly and is distinct from any controlled or uncontrolled tier. It is informative on its own: a high Not-Monitored proportion signals a follow-up process failure independent of disease control.

\subsubsection{Interpretation Guide}

\begin{itemize}[leftmargin=*]
    \item \textbf{High "Controlled" \%}: Effective disease management across the cluster's prevalent population.
    \item \textbf{High "Uncontrolled" \%}: Treatment intensification is needed; consider outreach to specific clusters or populations.
    \item \textbf{High "Not Monitored" \%}: Patients are diagnosed but not receiving guideline-recommended follow-up labs/visits. This is itself a care-gap signal and feeds into Reports 5--6.
    \item \textbf{Worst-level distribution}: Compare across clusters to identify where the most severely uncontrolled patients are concentrated.
\end{itemize}

% --------------------------------------------------------------------------
\subsection{Report 5 --- Care Gap (Quarterly)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
\textit{In each quarter of the report year, what proportion of prevalent patients completed a guideline-recommended follow-up visit with the relevant lab?}

\subsubsection{Measure Type}
\textbf{Quarterly care-gap completion rate} --- a process measure reflecting continuity of care for diagnosed patients.

\subsubsection{Base Population}
\textbf{Prevalent cohort as of the start of each quarter} (§3.1). The denominator is re-evaluated at each quarter boundary so that patients diagnosed mid-year enter the denominator in the quarter following their diagnosis.

\subsubsection{Quarterly Definitions}

\begin{table}[ht]
\centering
\caption{Quarter boundaries}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Quarter} & \textbf{Months} & \textbf{Reporting} \\
\midrule
Q1 & January -- March   & Months 1--3 \\
Q2 & April -- June      & Months 4--6 \\
Q3 & July -- September  & Months 7--9 \\
Q4 & October -- December & Months 10--12 \\
\bottomrule
\end{tabular}
\end{table}

A quarter is \textbf{completed} for a patient if the patient had $\geq 1$ follow-up marker in any month of that quarter (see condition-specific criteria in §7.0).

\subsubsection{Formula}

\begin{equation}
\text{Completion Rate}_{Q, \text{cluster}} =
\frac{\text{Prevalent patients completing follow-up in quarter } Q \text{ (cluster)}}
     {\text{Prevalent patients at start of quarter } Q \text{ (cluster)}} \times 100
\end{equation}

\subsubsection{Complementary Gap Rate}

The inverse metric is also reported directly:

\begin{equation}
\text{Gap Rate}_{Q, \text{cluster}} = 100\% - \text{Completion Rate}_{Q, \text{cluster}}
\end{equation}

\subsubsection{Interpretation Guide}

\begin{itemize}[leftmargin=*]
    \item \textbf{Declining completion across quarters}: Patients who completed follow-up early in the year are dropping off by year-end --- investigate clinic accessibility, scheduling, or reminder systems.
    \item \textbf{Uniformly low completion}: Systemic care-coordination failure; consider population-level outreach.
    \item \textbf{Cluster-level variation}: Identifies which facilities are performing well vs.\ under-performing on continuity-of-care for diagnosed patients.
\end{itemize}

% --------------------------------------------------------------------------
\subsection{Report 6 --- Care Gap (Annual)}
% --------------------------------------------------------------------------

\subsubsection{Epidemiological Question}
\textit{Across the four quarters of the report year, how many prevalent patients completed follow-up in 0, 1, 2, 3, or all 4 quarters --- and what proportion met the recommended target?}

\subsubsection{Measure Type}
\textbf{Annual care-gap distribution} --- an outcome-oriented summary of follow-up continuity across the year.

\subsubsection{Base Population}
\textbf{Prevalent cohort as of December 31} (§3.1). All prevalent patients are counted, regardless of when they were diagnosed.

\subsubsection{Formula}

For each prevalent patient, count the number of quarters in which they completed follow-up:

\begin{equation}
\text{quarters\_completed}_{\text{patient}} = \sum_{Q=1}^{4} \mathbb{1}[\text{patient completed follow-up in quarter } Q]
\end{equation}

Patients are then binned by $\text{quarters\_completed} \in \{0, 1, 2, 3, 4\}$. The aggregated report shows the count and percentage of prevalent patients in each bin, stratified by \texttt{health\_cluster}.

\subsubsection{Meeting Target}

The target is configured in \texttt{chi\_care\_gap\_config.target\_quarters\_completed} (default: 3). Patients are flagged:

\begin{equation}
\text{meets\_target}_{\text{patient}} = \mathbb{1}[\text{quarters\_completed}_{\text{patient}} \geq \text{target\_quarters\_completed}]
\end{equation}

The aggregated report includes a \texttt{$\geq$ Target} summary row per cluster, plus a grand-total row across all clusters:

\begin{equation}
\text{Pct Meeting Target}_{\text{cluster}} =
\frac{\text{Prevalent patients meeting target (cluster)}}
     {\text{Prevalent patients (cluster)}} \times 100
\end{equation}

\subsubsection{Interpretation Guide}

\begin{itemize}[leftmargin=*]
    \item \textbf{Most patients in 0--1 quarters}: Severe follow-up failure; the prevalent cohort is not receiving guideline-recommended care.
    \item \textbf{Bimodal distribution (0 and 4 quarters)}: Patients are either fully engaged or completely lost to follow-up; consider targeted outreach to the 0-quarter group.
    \item \textbf{Bell-shaped around 2 quarters}: Patients receive intermittent but not guideline-concordant care; consider structured recall systems.
    \item \textbf{Most patients at or above target}: Effective continuity of care for the prevalent population.
\end{itemize}

% --------------------------------------------------------------------------
\subsection{Control Thresholds}
% --------------------------------------------------------------------------

Control thresholds classify \textbf{prevalent} patients (Module 2 base population) into a controlled tier plus one or more uncontrolled tiers based on their most recent monitoring marker value. These are stored in the \texttt{CHI\_REPORTING.chi\_control\_thresholds} config table and are distinct from the screening thresholds in §5 (which classify at-risk patients for case-finding).

\subsubsection{Threshold Schema}

Each row of \texttt{chi\_control\_thresholds} defines one (condition, marker, gender, level) tuple with inclusive lower bound \texttt{min\_value} and exclusive upper bound \texttt{max\_value}:

\begin{table}[ht]
\centering\small
\caption{Control threshold schema (chi\_control\_thresholds)}
\begin{tabular}{@{}p{0.20\textwidth} p{0.13\textwidth} p{0.13\textwidth} p{0.10\textwidth} p{0.10\textwidth} p{0.10\textwidth} p{0.14\textwidth}@{}}
\toprule
\textbf{condition} & \textbf{marker} & \textbf{gender} & \textbf{min} & \textbf{max} & \textbf{order} & \textbf{label} \\
\midrule
\texttt{dm}  & \texttt{a1c}  & \texttt{All}    & \texttt{NULL} & 7.0  & 0 & \texttt{Controlled (A1C $<$ 7.0\%)} \\
\texttt{dm}  & \texttt{a1c}  & \texttt{All}    & 7.0  & 8.0  & 1 & \texttt{Uncontrolled (A1C 7.0--7.9\%)} \\
$\cdots$ & $\cdots$ & $\cdots$ & $\cdots$ & $\cdots$ & $\cdots$ & $\cdots$ \\
\bottomrule
\end{tabular}
\end{table}

A patient's marker value is matched against the row whose $[min, max)$ range contains it. \texttt{level\_order} is the sort key: 0 = best (controlled); higher values are worse. The \texttt{label} is the human-readable tier name shown in reports.

\subsubsection{Default Thresholds by Condition}

\paragraph{DM (A1C only).}

\begin{table}[ht]
\centering
\caption{DM control thresholds (A1C)}
\begin{tabular}{@{}cll@{}}
\toprule
\textbf{Order} & \textbf{Label} & \textbf{Range (\%)} \\
\midrule
0 & Controlled      & $< 7.0$ \\
1 & Uncontrolled    & $7.0$ -- $7.9$ \\
2 & Uncontrolled    & $8.0$ -- $8.9$ \\
3 & Uncontrolled    & $\geq 9.0$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{HTN (SYS + DIA, paired).}

\begin{table}[ht]
\centering
\caption{HTN control thresholds (SYS, DIA)}
\begin{tabular}{@{}clll@{}}
\toprule
\textbf{Marker} & \textbf{Order} & \textbf{Label} & \textbf{Range} \\
\midrule
SYS & 0 & Controlled      & $< 130$ mmHg \\
SYS & 1 & Uncontrolled    & $130$ -- $139$ \\
SYS & 2 & Uncontrolled    & $140$ -- $159$ \\
SYS & 3 & Uncontrolled    & $\geq 160$ \\
\midrule
DIA & 0 & Controlled      & $< 80$ mmHg \\
DIA & 1 & Uncontrolled    & $80$ -- $89$ \\
DIA & 2 & Uncontrolled    & $90$ -- $99$ \\
DIA & 3 & Uncontrolled    & $\geq 100$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{DLP (HDL, LDL, CHol, Trig --- $\max$ of 4).}

\begin{table}[ht]
\centering
\caption{DLP control thresholds (HDL, LDL, CHol, Trig)}
\begin{tabular}{@{}lllll@{}}
\toprule
\textbf{Marker} & \textbf{Gender} & \textbf{Order} & \textbf{Label} & \textbf{Range (mg/dL)} \\
\midrule
LDL  & All   & 0 & Controlled      & $< 100$ \\
LDL  & All   & 1 & Uncontrolled    & $100$ -- $129$ \\
LDL  & All   & 2 & Uncontrolled    & $130$ -- $159$ \\
LDL  & All   & 3 & Uncontrolled    & $\geq 160$ \\
\midrule
CHol & All   & 0 & Controlled      & $< 200$ \\
CHol & All   & 1 & Uncontrolled    & $200$ -- $239$ \\
CHol & All   & 2 & Uncontrolled    & $240$ -- $279$ \\
CHol & All   & 3 & Uncontrolled    & $\geq 280$ \\
\midrule
Trig & All   & 0 & Controlled      & $< 150$ \\
Trig & All   & 1 & Uncontrolled    & $150$ -- $199$ \\
Trig & All   & 2 & Uncontrolled    & $200$ -- $499$ \\
Trig & All   & 3 & Uncontrolled    & $\geq 500$ \\
\midrule
HDL  & Male  & 0 & Controlled      & $\geq 40$ \\
HDL  & Male  & 1 & Uncontrolled    & $< 40$ \\
HDL  & Female & 0 & Controlled     & $\geq 50$ \\
HDL  & Female & 1 & Uncontrolled   & $< 50$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{OB (BMI only, WHO).}

\begin{table}[ht]
\centering
\caption{OB control thresholds (BMI)}
\begin{tabular}{@{}cll@{}}
\toprule
\textbf{Order} & \textbf{Label} & \textbf{Range (kg/m$^2$)} \\
\midrule
0 & Controlled      & $18.5$ -- $24.9$ \\
1 & Uncontrolled    & $25.0$ -- $29.9$ \\
2 & Uncontrolled    & $30.0$ -- $34.9$ \\
3 & Uncontrolled    & $\geq 35.0$ \\
\bottomrule
\end{tabular}
\end{table}

% --------------------------------------------------------------------------
\subsection{Quarterly Definitions \& Follow-Up Criteria}
% --------------------------------------------------------------------------

The care-gap module partitions the report year into four calendar quarters and applies a condition-specific "completed follow-up" criterion per quarter. Both pieces are shared across Reports 5 and 6.

\subsubsection{Quarter Boundaries}

See Report 5 (§7.3) for the quarter-month mapping. Q1 = months 1--3, Q2 = 4--6, Q3 = 7--9, Q4 = 10--12. Boundaries align with calendar quarters.

\subsubsection{Follow-Up "Completed" Criteria}

A quarter is marked \textbf{completed} for a prevalent patient if any of the condition's follow-up markers was recorded in any month of that quarter:

\begin{table}[ht]
\centering
\caption{Follow-up "completed" criteria by condition (repeated from §7.0)}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Condition} & \textbf{Completed if patient had ...} & \textbf{Source table(s)} \\
\midrule
DM  & $\geq 1$ HbA1c measurement that quarter              & LAB + OBS \\
HTN & $\geq 1$ visit with paired SYS \textbf{and} DIA      & OBS only \\
DLP & $\geq 1$ lipid panel component (HDL/LDL/CHOL/TRIG)   & LAB + OBS \\
OB  & $\geq 1$ BMI measurement that quarter                 & OBS only \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Why HTN requires a paired visit.}
Screening classification already requires SYS+DIA to be from the same visit (§5.2) for clinical validity. The same constraint is carried into care-gap completion: a patient with only SYS or only DIA recorded that quarter is not counted as having completed follow-up.

\subsubsection{Target Configuration}

The quarterly target (default: 3 quarters) is stored in \texttt{chi\_care\_gap\_config.target\_quarters\_completed} and is read by Report 6 via \texttt{CROSS JOIN}. Changing the target does not require view re-creation.

% ============================================================================
\section{Assumptions \& Limitations}
% ============================================================================

\subsection{Shared Assumptions (Module 1 \& Module 2)}

\begin{enumerate}[leftmargin=*, label=\textbf{SA\arabic*}]
    \item \textbf{Complete diagnosis capture}. It is assumed that all diagnoses are recorded in the EMR with accurate ICD-10 codes and dates. Undiagnosed patients are misclassified as at-risk (Module 1) or non-prevalent (Module 2).
    \item \textbf{Lab result accuracy}. Both screening classifications (Module 1) and control classifications (Module 2) assume lab/observation values are correctly recorded. Transcription errors may cause misclassification. The DM dual-unit auto-detection mitigates but does not eliminate this.
    \item \textbf{PHC assignment completeness}. Patients without a PHC record are assumed to be genuinely unassigned, not the result of a data integration gap.
    \item \textbf{Visit-date alignment}. Lab results and observations are attributed to the month of the associated patient visit, not the date the result was entered or verified.
    \item \textbf{No loss to follow-up}. Patients who leave the catchment area but have no death record are assumed to remain in the population. This may overstate the denominator for both modules.
\end{enumerate}

\subsection{Module 1 Assumptions \& Limitations}

\begin{enumerate}[leftmargin=*, label=\textbf{M1-L\arabic*}]
    \item \textbf{Screening $\neq$ diagnosis}. A patient with an abnormal screening result does not necessarily have the disease. The screening report measures test coverage and results, not confirmed diagnoses.
    \item \textbf{First diagnosis $\neq$ disease onset}. The incidence measure captures the date a diagnosis was \textit{recorded}, not the date the disease \textit{developed}. A patient diagnosed with E11 in June may have had undiagnosed diabetes for years.
    \item \textbf{Screening encounter double-counting}. The annual cumulative screening rate counts screening \textit{events}, not unique patients. A patient screened every month contributes 12 to the numerator. This metric measures testing volume, not the proportion of the population ever screened.
    \item \textbf{Denominator stability in small clusters}. For small health clusters, a single incident case can produce a very high rate (e.g., 1 case / 2 at-risk = 50,000 per 100k). Rates for clusters with $< 5$ at-risk patients should be interpreted with caution.
    \item \textbf{ICD-10 coding variability}. Different clinicians may code the same condition differently. A patient with Type 2 diabetes coded as E14 (unspecified DM) would be counted in the prevalent cohort for screening/incidence exclusion but not in the E11-specific prevalence numerator.
    \item \textbf{Cross-sectional prevalence denominator}. The prevalence denominator is fixed at January 1. Patients who turn 18, receive a National ID, or immigrate into the catchment during the year are not captured.
\end{enumerate}

\subsection{Module 2 Assumptions \& Limitations}

\begin{enumerate}[leftmargin=*, label=\textbf{M2-L\arabic*}]
    \item \textbf{Control thresholds are treatment targets, not physiological norms}. The default A1C target ($< 7.0\%$), BP target ($< 130 / < 80$), and lipid targets reflect guideline-recommended treatment goals for the general diabetic / hypertensive / dyslipidemic population. Individualized targets may differ (e.g., tighter control for some patients, more relaxed for frail elderly). The report applies a uniform threshold across all prevalent patients; clinical interpretation should account for individual variation.
    \item \textbf{``Completed'' is a process proxy, not an outcome}. Report 5 measures whether a lab/visit was \textit{recorded}, not whether the patient's disease is \textit{well managed}. A patient with quarterly HbA1c measurements all above 9.0\% is counted as having completed follow-up each quarter --- but is severely uncontrolled. Combine Report 4 (control level) and Report 5 (completion) for a complete picture.
    \item \textbf{Mid-year diagnoses contribute partial-year data}. A patient diagnosed in October contributes to the prevalent cohort as of December 31 (Report 4 and Report 6 denominators) and to Report 5 from Q4 onward only. They have 1 quarter to complete follow-up, making them structurally unable to meet a target of 3 quarters. Report 6's annual distribution should be interpreted with awareness of the incident case mix.
    \item \textbf{``Not Monitored'' conflates care gaps and data gaps}. A prevalent patient with no measurement during the year may genuinely be lost to follow-up (a care-gap signal) or may have measurements recorded in systems not captured by the EMR (a data-coverage issue). Report 5's per-quarter completion rate is the primary mechanism to distinguish these.
    \item \textbf{Multi-marker aggregation masks individual marker issues}. For HTN (SYS+DIA) and DLP (4 markers), the overall control level is the $\max$ across markers. A patient with controlled LDL but severely uncontrolled HDL is reported at the worst marker's level only --- the underlying multi-marker profile is not visible in the aggregated report.
    \item \textbf{HDL gender-specific thresholds assume recorded gender is correct}. Patients with miscoded or missing gender inherit Male thresholds by default. Gender-specific reports should be cross-checked against the cohort view's gender distribution.
    \item \textbf{Quarterly boundaries align with calendar quarters, not patient-level follow-up windows}. Q1 = Jan--Mar is a calendar convention. A patient diagnosed on March 15 has only 16 days to complete Q1 follow-up before the quarter ends; the report does not normalize for time-in-quarter.
\end{enumerate}

\subsection{Prediabetes-Specific Limitations}
These apply to the \textbf{Prediabetes-specific} Module-1 reports (Prediabetes Annual Prevalence [Report 2] and Prediabetes Monthly Incidence [Report 3]):
\begin{enumerate}[leftmargin=*]
    \item \textbf{R73.03 / R73.* scope is locked to R73.03}. Other \texttt{R73.*} codes (e.g., \texttt{R73.02} impaired glucose tolerance; \texttt{R73.09} other abnormal glucose) are \emph{not} counted in the prediabetes at-risk / prevalent cohorts. If broader inclusion becomes clinically warranted, the \texttt{WHERE} clauses in \texttt{stg\_prediab\_cohort.prediab\_diagnosis\_summary} and \texttt{stg\_prediab\_diagnosis} are the only changes required.
    \item \textbf{Prediabetes at-risk pool excludes R73.03 patients (inverse of DM)}. The DM at-risk pool \emph{includes} patients with R73.03 (they can be screened and become incident E11 cases); the Prediabetes at-risk pool \emph{excludes} them (they're prevalent). The two cohorts are deliberately inverse --- see §3.2.
    \item \textbf{No prediabetes-specific screening lab} is captured in the EMR's standard observation set. FBS and A1C exist as \emph{DM} screening markers; they are not modeled as prediabetes screening markers here. As a result, there is \textbf{no Module-1 Screening Report for Prediabetes} --- only Prevalence (annual) and Incidence (monthly).
    \item \textbf{Risk-factor flags (BMI latest, HTN dx, DLP dx, family-history placeholder, GDM, PCOS) are computed in \texttt{stg\_prediab\_cohort}}. They are needed by both the old \texttt{is\_high\_risk\_prediab} column on that view and the generic High-Risk Patients report. Keeping them on the cohort view (pred-iabetes-specific) avoids recomputing per month.
\end{enumerate}

\subsection{Generic High-Risk Limitations}
These apply to the \textbf{generic} Module-2 High-Risk Patients report (Report 7):
\begin{enumerate}[leftmargin=*]
    \item \textbf{v1 covers only Prediabetes}. For DM, HTN, DLP, and OB the \texttt{chi\_high\_risk\_factors} config table has no rows, so the report produces no output for those conditions. Extending to a new condition requires (a) \texttt{INSERT} rows into \texttt{chi\_high\_risk\_factors} with the appropriate \texttt{source\_view} / \texttt{source\_column} pairs, and (b) extending the \texttt{CASE} chain in \texttt{stg\_high\_risk\_patient.factor\_evaluations} to recognize the new \texttt{source\_column} values.
    \item \textbf{Family-history placeholder wired as \texttt{always\_false} sentinel}. The \texttt{chi\_high\_risk\_factors.source\_column = 'always\_false'} pattern returns \texttt{FALSE} for every patient --- the report treats this factor as ``not yet implemented.'' When a real source exists (e.g., a \texttt{FAMILY\_HISTORY} table with ICD-10 Z83.3), the sentinel row is replaced by a row pointing at the real view/column. No SQL change in \texttt{stg\_high\_risk\_patient} is needed beyond ensuring the new \texttt{source\_column} is named in the CASE chain.
    \item \textbf{Minimum factor threshold ($N = 2$) is hardcoded} in \texttt{stg\_high\_risk\_patient}. A future extension could add a \texttt{min\_factors\_for\_high\_risk} column to \texttt{chi\_high\_risk\_factors} (per-condition) and parameterize the view. For v1 the constant is fine --- every condition in the foreseeable future will use 2.
    \item \textbf{Per-condition source\_view joins are inlined in the view's CASE chain} (not driven by the config table's \texttt{source\_view} value). This is a deliberate v1 simplification: \texttt{source\_view} and \texttt{source\_column} in the config are documentation/audit labels, but the actual JOIN logic lives in \texttt{stg\_high\_risk\_patient}. A future extension could replace the CASE chain with a stored procedure that does dynamic SQL driven by the config.
    \item \textbf{Conditions with 0 factors produce no rows} in \texttt{rpt\_high\_risk\_patients\_annual} (not zero-filled placeholder rows). Reading the report requires the reader to know which conditions are active --- by design, an empty section is a strong signal that no risk factors have been configured yet.
    \item \textbf{The report is silent when the patient's prevalent cohort is empty} (e.g., zero OB-prevalent patients in a cluster). This is correct behaviour; the row simply doesn't appear in the output.
\end{enumerate}

% ============================================================================
\section{Reference: Formula Index}
% ============================================================================

\begin{table}[ht]
\centering\small
\caption{Summary of all rate formulas (10 reports across 2 modules)}
\begin{tabular}{@{}p{0.10\textwidth} p{0.16\textwidth} p{0.34\textwidth} p{0.34\textwidth}@{}}
\toprule
\textbf{Module} & \textbf{Report} & \textbf{Detail Row Formula} & \textbf{Subtotal / Grand Total Formula} \\
\midrule
\multirow{5}{*}{M1} &
Screening &
$\dfrac{\text{Screened}_m}{\text{At-Risk}_m} \times 100$ &
$\dfrac{\sum\text{Screened}_m}{\sum\text{At-Risk}_m} \times 100$ (cumulative) \\[14pt]
& Prevalence &
--- (annual only) &
$\dfrac{\text{Prevalent}_{\text{Dec 31}}}{\text{Total Pop}_{\text{Jan 1}}} \times 100$ \\[14pt]
& Incidence &
$\dfrac{\text{New Cases}_m}{\text{At-Risk}_m} \times 100{,}000$ &
$\dfrac{\sum\text{New Cases}_m}{\text{At-Risk}_{\text{Jan}}} \times 100{,}000$ (annualized) \\[10pt]
& Prediabetes Prevalence &
--- (annual only) &
$\dfrac{\text{R73-prevalent}_{\text{Dec 31}}}{\text{Total Pop}_{\text{Jan 1}}} \times 100$ \\[14pt]
& Prediabetes Incidence &
$\dfrac{\text{New R73.03}_m}{\text{At-risk prediab}_m} \times 100{,}000$ &
$\dfrac{\sum\text{New R73.03}_m}{\text{At-risk prediab}_{\text{Jan}}} \times 100{,}000$ (annualized) \\[10pt]
\midrule
\multirow{4}{*}{M2} &
High-Risk Patients &
--- (annual only) &
$\dfrac{\text{Prevalent patients with risk\_factor\_count} \geq 2}{\text{Prevalent patients (per condition)}} \times 100$ \\[14pt]
& Control &
$\dfrac{\text{Pts in level}_{c}}{\text{Prevalent}_{c}} \times 100$ &
$\dfrac{\sum\text{Pts in level}}{\sum\text{Prevalent}} \times 100$ \\[14pt]
& Care Gap Q &
$\dfrac{\text{Completed}_Q}{\text{Prevalent}_Q} \times 100$ &
$\dfrac{\sum\text{Completed}_Q}{\sum\text{Prevalent}_Q} \times 100$ \\[14pt]
& Care Gap Annual &
$\dfrac{\text{Pts in bin}}{\text{Prevalent}} \times 100$ &
Bins + ``$\geq$ Target'' row per cluster \\[4pt]
\bottomrule
\end{tabular}
\end{table}

Where $m$ = month (1--12), $c$ = cluster, $Q$ = quarter (1--4), ``bin'' = $\{0, 1, 2, 3, 4\}$ quarters completed.

\end{document}